dispRity manual

### 2017-06-24 doc/dispRity-manual.Rmd
updates (tests) to manual

You can install this package easily if you use the latest version of `R` (> 3.0.0) and `devtools`.

@@

## Creating time slices subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
## Bootstrapping on the time binning/slicing subsamples
boot.matrix(time_slices, bootstraps = 1000)
```

@@

```{r}
## Example data from previous sections
crown_stem <- custom.subsamples(BeckLee_mat50,
    group = list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44)))
## Bootstrapping and rarefying these groups
boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)
## Calculate disparity
disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))

## Creating time slices subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
## Bootstrapping on the time binning/slicing subsamples
boot_time_slices <- boot.matrix(time_slices, bootstraps = 1000)
## Calculate disparity
disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))
```

@@

```{r, fig.width=8, fig.height=8}
## Graphical options
par(bty = "n")

@@

```{r, fig.width=8, fig.height=8}
## Graphical options
par(bty = "n")

## Plotting the continuous disparity with a fixed y axis
plot(disparity_time_slices, ylim = c(1, 4))
## Adding the discrete data
plot(disparity_crown_stem, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
    add = TRUE)

@@

<!-- Change this example to add time bins! -->

@@

Finally, the `dispRity` package allows to apply some test functions to the calculated disparity in order to test hypothesis.
The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object, a `test` and a `comparisons` argument.

@@

It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsamples names to create personalised test.
Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` and the `dispRity::sequential.test` have a specific way to be applied to the data and therefore ignore the `comparisons` argument.

@@

```{r, eval=FALSE}
## Performing a t-test to the test a difference in disparity between crown and stem mammals
test.dispRity(disp_crown_stemBS, test = t.test)

## Performing the same test but with the detailed t.test output
test.dispRity(disp_crown_stemBS, test = t.test, details = TRUE)

## Performing a Wilcoxon test applied to sliced disparity the data sequentially with correction
test.dispRity(disp_time_slicesBS, test = wilcox.test, comparisons = "sequential",
    correction = "bonferroni")

## Measuring the overlap between distributions in the time bins (using the implemented
## Bhattacharyya Coefficient function - see ?bhatt.coeff)
test.dispRity(disp_time_binsBS, test = bhatt.coeff)
```

@@

```{r, eval=FALSE}
## Performing and ANOVA applied to the same data
slice_aov <- (test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all"))

## The output is a regular ANOVA output...
class(slice_aov)

## That can be summarised using summary
summary(slice_aov)
```

@@

@@

```{r, eval=FALSE}
## Sorting the disparity subsamples in inverse alphabetic order
summary(sort(disp_time_slicesBS, decreasing = TRUE))

>>>>

You can install this package easily if you use the latest version of `R` (> 3.3.0) and `devtools`.

@@

## Creating time slices subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
## Bootstrapping on the time binning/slicing subsamples
boot.matrix(time_slices, bootstraps = 100)
```

@@

```{r}
## Example data from previous sections
crown_stem <- custom.subsamples(BeckLee_mat50,
    group = list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44)))
## Bootstrapping and rarefying these groups
boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
## Calculate disparity
disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))

## Creating time slices subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
## Bootstrapping on the time binning/slicing subsamples
boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
## Calculate disparity
disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))

## Creating time bins subsamples
time_bins <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
## Bootstrapping on the time binning/slicing subsamples
boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
## Calculate disparity
disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
```

@@

```{r, fig.width=8, fig.height=8}
## Graphical options
op <- par(bty = "n")

@@

```{r, fig.width=8, fig.height=8}
## Graphical options
op <- par(bty = "n")

## Plotting the continuous disparity with a fixed y axis
plot(disparity_time_slices, ylim = c(1, 4))
## Adding the discrete data
plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
    add = TRUE)

@@

@@

The `dispRity` package allows to apply some test functions to the calculated disparity in order to test hypothesis.
The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object, a `test` and a `comparisons` argument.

@@

It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsamples names to create personalised test.
Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` that has a specific way to be applied to the data and therefore ignore the `comparisons` argument. 
<!-- Add sequential test one day! -->

@@

```{r}
## Performing a t-test to the test a difference in disparity between crown and stem mammals
test.dispRity(disparity_crown_stem, test = t.test)

## Performing the same test but with the detailed t.test output
test.dispRity(disparity_crown_stem, test = t.test, details = TRUE)

## Performing a Wilcoxon test applied to sliced disparity the data sequentially with correction
test.dispRity(disparity_time_slices, test = wilcox.test, comparisons = "sequential",
    correction = "bonferroni")

## Measuring the overlap between distributions in the time bins (using the implemented
## Bhattacharyya Coefficient function - see ?bhatt.coeff)
test.dispRity(disparity_time_bins, test = bhatt.coeff)
```

@@

```{r}
## Performing and ANOVA applied to the same data
slice_aov <- test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
slice_aov

## The output is a regular ANOVA output
class(slice_aov)

## That can be summarised using summary
summary(slice_aov)
```
Of course, following the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
The only condition is that the test can be applied to at least two distributions.
In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `aov` for example).

@@

Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
This value can be of course pseudo-replicated (boostraped) in order to obtain a more confident distribution of this summary metric.
However, another way to perform disparity analysis is to use the whole distribution rather than their summary (e.g. the variances or the ranges).

This is possible in the `dispRity` package by calculating disparity as a level 2 metric only!
Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxa and their centroid as disparity

```{r}
## Measuring disparity as a whole distribution
disparity_centroids <- dispRity(boot_time_slices, metric = centroids)
```

The resulting disparity object being a level 2, it can easily also be transformed into a level 1 one by for example measuring the median distance of all these distributions:

```{r}
## Measuring median disparity in each time slice
disparity_centroids_median <- dispRity(disparity_centroids, metric = median)
```

And we can now compare the differences between both methods:

```{r}
## Summarising both disparity measurements:
## The distributions:
summary(disparity_centroids)
## The summary of the distributions (as median)
summary(disparity_centroids_median)
```

We can see that the summary message for the distribution is slightly different than before.
Here `summary` also displays the observed central tendency (i.e. the central tendency of the measured distributions).
Note that, as expected, it is the same in both metrics!

<!-- STOPPED HERE -->

```{r, fig.width=8, fig.height=8}
## Graphical parameters
op <- par(bty = "n", mfrow = c(1,2))

## Plotting both disparity measurements
plot(disparity_centroids, ylab = "Distribution of all the distances")
plot(disparity_centroids_median, ylab = "Distribution of the medians of all the distances")

par(op)
```

The resulted distributions can then be studied the same way as described above.

@@

```{r, eval=FALSE}
## Sorting the disparity subsamples in inverse alphabetic order
summary(sort(disp_time_slicesBS, decreasing = TRUE))
```
